---
title: "Assignment 3"
author: "Erika Yazmin Blanco, Felipe Manzi, Olsa Reçica, Olta Berani"
date: "2026-03-03"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Task 1

```{r cars}
library(terra)
library(sf)
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
```

```{r pressure, echo=FALSE}
# Load SPEI (NetCDF)
spei <- rast("spei01.nc")

# IMPORTANT: assign a proper monthly time index (SPEIbase is typically 1901-present)
time(spei) <- seq.Date(from = as.Date("1901-01-01"), by = "month", length.out = nlyr(spei))

# Crop to US bounding box (speeds up)
spei_us <- crop(spei, ext(-125, -66, 24, 50))

# Filter to last 50 years (edit as needed)
spei_us <- spei_us[[ time(spei_us) >= as.Date("1975-01-01") & time(spei_us) <= as.Date("2025-12-01") ]]

# Rename layers to dates so pivoting is robust
names(spei_us) <- format(time(spei_us), "%Y-%m-%d")
```

```{r}
states <- st_read("cb_2018_us_state_20m.shp", quiet = TRUE) |>
  filter(!STUSPS %in% c("AK", "HI", "PR")) |>
  st_transform(crs(spei_us))
```


```{r}
region_map <- c(
  # Northeast
  CT="Northeast", ME="Northeast", MA="Northeast", NH="Northeast", RI="Northeast", VT="Northeast",
  NJ="Northeast", NY="Northeast", PA="Northeast",
  # Midwest
  IL="Midwest", IN="Midwest", MI="Midwest", OH="Midwest", WI="Midwest",
  IA="Midwest", KS="Midwest", MN="Midwest", MO="Midwest", NE="Midwest", ND="Midwest", SD="Midwest",
  # South
  DE="South", FL="South", GA="South", MD="South", NC="South", SC="South", VA="South", WV="South", DC="South",
  AL="South", KY="South", MS="South", TN="South",
  AR="South", LA="South", OK="South", TX="South",
  # West (CONUS)
  AZ="West", CO="West", ID="West", MT="West", NV="West", NM="West", UT="West", WY="West",
  CA="West", OR="West", WA="West"
)

states$region <- unname(region_map[states$STUSPS])

regions <- states |>
  filter(!is.na(region)) |>
  group_by(region) |>
  summarise(geometry = st_union(geometry), .groups = "drop")

regions_v <- vect(regions)
```


```{r}
# Area-weighted mean SPEI by region
reg_ex <- terra::extract(spei_us, regions_v, fun = mean, weights = TRUE, na.rm = TRUE)

# Build panel: region x date x spei
panel <- reg_ex |>
  as.data.frame() |>
  mutate(region = regions$region[ID]) |>
  pivot_longer(
    cols = -c(ID, region),
    names_to = "date",
    values_to = "spei"
  ) |>
  mutate(date = as.Date(date)) |>
  select(region, date, spei) |>
  arrange(region, date)

# Yearly SPEI per region
panel_year <- panel |>
  mutate(year = year(date)) |>
  group_by(region, year) |>
  summarise(spei = mean(spei, na.rm = TRUE), .groups = "drop")
```

```{r}
avg_year <- panel_year |>
  group_by(year) |>
  summarise(us_mean = mean(spei, na.rm = TRUE), .groups = "drop")
```

```{r}
ggplot() +

  # Regional yearly lines (no smooth, no CI)
  geom_line(
    data = panel_year,
    aes(x = year, y = spei, color = region, group = region),
    alpha = 0.6,
    linewidth = 0.8
  ) +


  # US average smooth WITH confidence interval ribbon (transparent)
  geom_smooth(
    data = avg_year,
    aes(x = year, y = us_mean),
    method = "loess",        # or "gam" if you prefer
    se = TRUE,
    color = "black",
    fill = "black",
    alpha = 0.15,
    linewidth = 1.3
  ) +

  labs(
    x = "Year",
    y = "Yearly SPEI",
    color = "Region",
    title = "Yearly SPEI by Region with Smoothed US Average (95% CI)"
  ) +
  theme_minimal()
```

# Task 2

```{r libraries}
library(sf)
library(terra)
library(tidyverse)
library(gdistance)
library(raster)
```


1. We load the ne_10m_populated_places shapefile and filter to Spain. 
We then rank cities by population and keep the top 10.
```{r}
# load country boundaries
countries <- st_read("data/ne_10m_admin_0_countries.shp")
# extract Spain polygon
spain <- countries %>% filter(ADMIN == "Spain")
# load populated places
places <- st_read("data/ne_10m_populated_places.shp")
# make sure CRS matches
places <- st_transform(places, st_crs(spain))
# filter places within Spain
places_spain <- st_filter(places, spain)
# get top 10 by population
top10 <- places_spain %>%
  arrange(desc(POP_MAX)) %>%
  slice(1:10)
print(top10 %>% st_drop_geometry() %>% dplyr::select(NAME, POP_MAX))
# plot
ggplot() +
  geom_sf(data = spain) +
  geom_sf(data = top10 %>% filter(!NAME %in% c("Madrid", "Vigo")), 
          color = "gray40", size = 3) +
  geom_sf(data = top10 %>% filter(NAME %in% c("Madrid", "Vigo")),
          aes(color = NAME), size = 3) +
  scale_color_manual(values = c("Madrid" = "#e63946", "Vigo" = "#2a9d8f")) +
  labs(title = "Top 10 Populated Places in Spain", color = "NAME") +
  theme_bw() +
coord_sf(xlim = c(-10, 5), ylim = c(35.5, 44.5))
```

2. We load the gROADS-v1-europe road network and crop it 
to Spain's boundary using st_intersection().
```{r}
# load gROADS Europe shapefile
roads_europe <- st_read("data/gROADS-v1-europe.shp")

# reproject to match Spain
roads_europe <- st_transform(roads_europe, st_crs(spain))

# crop roads to Spain
roads_spain <- st_intersection(roads_europe, spain)

# plot
ggplot() +
  geom_sf(data = spain) +
  geom_sf(data = roads_spain, linewidth = 0.2, color = "gray30") +
  geom_sf(data = top10, color = "red", size = 2) +
  labs(title = "Road Network in Spain + Top 10 Cities") +
  theme_bw() +
coord_sf(xlim = c(-10, 5), ylim = c(35.5, 44.5))
```

3. We build a friction surface by rasterizing the road network. 
Road cells get a cost of 1 and off-road cells get a cost of 100. 
We then build a transition matrix and compute the full 10x10 
bilateral distance matrix between all city pairs.
```{r}
# convert Spain to terra vector
spain_vect <- vect(spain)

# create raster template over Spain at 0.05 degree resolution
r_template <- rast(spain_vect, res = 0.05)

# rasterize roads: cells with roads = 1
roads_rast <- rasterize(vect(roads_spain), r_template, field = 1)

# assign friction: 1 on-road, 100 off-road
fv <- values(roads_rast)
fv[is.na(fv)] <- 100
fv[fv == 1]   <- 1
values(roads_rast) <- fv

# mask to Spain only
spain_mask <- rasterize(vect(spain), r_template, field = 1)
roads_rast <- mask(roads_rast, spain_mask)

# plot
ext_spain <- c(-9.5, 4.5, 36, 44)
plot(roads_rast, main = "Friction Surface (1 = road, 100 = off-road)", ext = ext_spain)
plot(spain_vect, add = TRUE)
```

```{r}
# convert to old raster format (required by gdistance)
r_raster <- raster::raster(roads_rast)

# build transition matrix (8 directions, mean function)
tr_matrix <- transition(
  x                  = r_raster,
  transitionFunction = mean,
  directions         = 8
)

# geocorrect for diagonal distances
tr_matrix <- geoCorrection(tr_matrix, type = "c")
```

```{r}
# compute full 10x10 cost-distance matrix
dist_friction <- costDistance(
  tr_matrix,
  as(top10, "Spatial")
) %>% as.matrix()

rownames(dist_friction) <- top10$NAME
colnames(dist_friction) <- top10$NAME

# display matrix
print(round(dist_friction, 0))
```

4. We compare cost distances from Madrid and Vigo to all other 
cities using geom_density(). The city with the higher mean 
distance is more isolated.

```{r}
# extract distances from Madrid and Vigo to all other cities
madrid_idx <- which(rownames(dist_friction) == "Madrid")
vigo_idx   <- which(rownames(dist_friction) == "Vigo")

dist_df <- bind_rows(
  tibble(city = "Madrid", distance = dist_friction[madrid_idx, -madrid_idx]),
  tibble(city = "Vigo",   distance = dist_friction[vigo_idx,   -vigo_idx])
)

# dynamic x-axis upper limit
max_val <- ceiling(max(dist_df$distance) / 100) * 100
# density plot
ggplot(dist_df, aes(x = distance, fill = city, color = city)) +
  geom_density(alpha = 0.4, linewidth = 1) +
  scale_fill_manual(values = c("Madrid" = "#e63946", 
                               "Vigo"   = "#2a9d8f")) +
  scale_color_manual(values = c("Madrid" = "#e63946", 
                                "Vigo"   = "#2a9d8f")) +
  scale_x_continuous(
    limits = c(0, max_val),
    breaks = seq(0, max_val, by = 1800),
    expand = c(0, 0)
  ) +
  labs(
    title    = "Transportation Isolation: Madrid vs. Vigo",
    subtitle = "Friction-surface distances to the other top-10 cities",
    x        = "Cost Distance",
    y        = "Density",
    fill     = "Origin City",
    color    = "Origin City"
  ) +
  theme_bw()

# conclusion
m_madrid <- round(mean(dist_friction[madrid_idx, -madrid_idx]), 1)
m_vigo   <- round(mean(dist_friction[vigo_idx, -vigo_idx]), 1)

cat("Mean distance from Madrid:", m_madrid, "\n")
cat("Mean distance from Vigo:", m_vigo, "\n\n")
cat("Vigo has a higher average friction-surface distance to the other top-10 cities,\n",
    "indicating greater transportation isolation relative to Madrid.")
```

