---
title: "Take-home Assignment (2/2): Transportation Centrality and Isolation"
Task: 2
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r}
library(sf)
library(terra)
library(tidyverse)
library(gdistance)
library(raster)
```


1. We load the ne_10m_populated_places shapefile and filter to Spain. 
We then rank cities by population and keep the top 10.
```{r}
# load country boundaries
countries <- st_read("data/ne_10m_admin_0_countries.shp")
# extract Spain polygon
spain <- countries %>% filter(ADMIN == "Spain")
# load populated places
places <- st_read("data/ne_10m_populated_places.shp")
# make sure CRS matches
places <- st_transform(places, st_crs(spain))
# filter places within Spain
places_spain <- st_filter(places, spain)
# get top 10 by population
top10 <- places_spain %>%
  arrange(desc(POP_MAX)) %>%
  slice(1:10)
print(top10 %>% st_drop_geometry() %>% dplyr::select(NAME, POP_MAX))
# plot
ggplot() +
  geom_sf(data = spain) +
  geom_sf(data = top10 %>% filter(!NAME %in% c("Madrid", "Vigo")), 
          color = "gray40", size = 3) +
  geom_sf(data = top10 %>% filter(NAME %in% c("Madrid", "Vigo")),
          aes(color = NAME), size = 3) +
  scale_color_manual(values = c("Madrid" = "#e63946", "Vigo" = "#2a9d8f")) +
  labs(title = "Top 10 Populated Places in Spain", color = "NAME") +
  theme_bw() +
coord_sf(xlim = c(-10, 5), ylim = c(35.5, 44.5))
```

2. We load the gROADS-v1-europe road network and crop it 
to Spain's boundary using st_intersection().
```{r}
# load gROADS Europe shapefile
roads_europe <- st_read("data/gROADS-v1-europe.shp")

# reproject to match Spain
roads_europe <- st_transform(roads_europe, st_crs(spain))

# crop roads to Spain
roads_spain <- st_intersection(roads_europe, spain)

# plot
ggplot() +
  geom_sf(data = spain) +
  geom_sf(data = roads_spain, linewidth = 0.2, color = "gray30") +
  geom_sf(data = top10, color = "red", size = 2) +
  labs(title = "Road Network in Spain + Top 10 Cities") +
  theme_bw() +
coord_sf(xlim = c(-10, 5), ylim = c(35.5, 44.5))
```
3. We build a friction surface by rasterizing the road network. 
Road cells get a cost of 1 and off-road cells get a cost of 100. 
We then build a transition matrix and compute the full 10x10 
bilateral distance matrix between all city pairs.
```{r}
# create raster template over Spain at 0.05 degree resolution
r_template <- rast(ext(spain), res = 0.05, crs = "EPSG:4326")

# rasterize roads: cells with roads = 1
roads_rast <- rasterize(vect(roads_spain), r_template, field = 1)

# assign friction: 1 on-road, 100 off-road
fv <- values(roads_rast)
fv[is.na(fv)] <- 100
fv[fv == 1]   <- 1
values(roads_rast) <- fv

# mask to Spain only
spain_mask <- rasterize(vect(spain), r_template, field = 1)
roads_rast <- mask(roads_rast, spain_mask)

# plot
ext_spain <- c(-9.5, 4.5, 36, 44)
plot(roads_rast, main = "Friction Surface (1 = road, 100 = off-road)", ext = ext_spain)
plot(vect(spain), add = TRUE)
```

```{r}
# convert to old raster format (required by gdistance)
r_raster <- raster::raster(roads_rast)

# build transition matrix (8 directions, mean function)
tr_matrix <- transition(
  x                  = r_raster,
  transitionFunction = mean,
  directions         = 8
)

# geocorrect for diagonal distances
tr_matrix <- geoCorrection(tr_matrix, type = "c")
```

```{r}
# compute full 10x10 cost-distance matrix
dist_friction <- costDistance(
  tr_matrix,
  as(top10, "Spatial")
) %>% as.matrix()

rownames(dist_friction) <- top10$NAME
colnames(dist_friction) <- top10$NAME

# display matrix
print(round(dist_friction, 0))
```
4. We compare cost distances from Madrid and Vigo to all other 
cities using geom_density(). The city with the higher mean 
distance is more isolated.

```{r}
# extract distances from Madrid and Vigo to all other cities
madrid_idx <- which(rownames(dist_friction) == "Madrid")
vigo_idx   <- which(rownames(dist_friction) == "Vigo")

dist_df <- bind_rows(
  tibble(city = "Madrid", distance = dist_friction[madrid_idx, -madrid_idx]),
  tibble(city = "Vigo",   distance = dist_friction[vigo_idx,   -vigo_idx])
)

# density plot
ggplot(dist_df, aes(x = distance, fill = city, color = city)) +
  geom_density(alpha = 0.4, linewidth = 1) +
  scale_fill_manual(values  = c("Madrid" = "#e63946", "Vigo" = "#2a9d8f")) +
  scale_color_manual(values = c("Madrid" = "#e63946", "Vigo" = "#2a9d8f")) +
  labs(
    title    = "Transportation Isolation: Madrid vs. Vigo",
    subtitle = "Friction-surface distances to the other top-10 cities",
    x        = "Cost Distance",
    y        = "Density",
    fill     = "Origin City",
    color    = "Origin City"
  ) +
  theme_bw()

# conclusion
cat("Mean distance from Madrid:", round(mean(dist_friction[madrid_idx, -madrid_idx]), 1),
    "\nMean distance from Vigo:  ", round(mean(dist_friction[vigo_idx, -vigo_idx]), 1),
    "\n\nVigo has a higher mean distance, meaning it is more isolated than Madrid.")
```