---
title: "Baseline Maps: Oil Infrastructure & Armed Conflict in the Middle East"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

We start by loading the cleaned datasets we prepared earlier  conflict events from ACLED, country boundaries, oil pipelines from GEM, and extraction sites from the Global Oil & Gas Production Tracker. All layers are in CRS 4326 so they share the same coordinate reference system.

```{r load}
library(sf)
library(ggplot2)
library(ggdensity)

conflict_list  <- readRDS("data/clean/acled_clean.rds")
conflict       <- conflict_list$crs_4326

countries_list <- readRDS("data/clean/countries_clean.rds")
countries      <- countries_list$crs_4326

gem   <- readRDS("data/clean/gem_clean.rds")
gogpt <- readRDS("data/clean/gogpt_clean.rds")
```

Our first map plots every armed conflict event from ACLED as a red point over the eight study countries. This gives us a quick baseline picture of where violence is concentrated across the region.

```{r map1}
legend_data1 <- data.frame(
  x = c(0),
  y = c(0),
  type = c("Conflict Events")
)

ggplot() +
  geom_sf(data = countries, fill = "#f5f0e8", color = "grey40", linewidth = 0.4) +
  geom_sf(data = conflict, color = "#e8000d", size = 0.6, alpha = 0.7) +
  geom_point(data = legend_data1, aes(x = x, y = y, color = type), size = 0, alpha = 0) +
  geom_sf_text(data = countries, aes(label = ADMIN), size = 2.5, color = "grey30", fontface = "italic") +
  scale_color_manual(
    name   = NULL,
    values = c("Conflict Events" = "#e8000d"),
    guide  = guide_legend(override.aes = list(size = 3, alpha = 1))
  ) +
  coord_sf(xlim = c(34, 62), ylim = c(12, 42), expand = FALSE) +
  labs(
    title    = "Armed Conflict Events in the Middle East",
    subtitle = "Based on ACLED data across eight study countries",
    caption  = "Source: Armed Conflict Location & Event Data (ACLED)"
  ) +
  theme_void() +
  theme(
    plot.title      = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle   = element_text(size = 10, color = "grey40", hjust = 0.5),
    plot.caption    = element_text(size = 8,  color = "grey50", hjust = 0.5),
    plot.margin     = margin(10, 10, 10, 10),
    legend.position = "left",
    legend.text     = element_text(size = 8)
  )
```



Next, we mapped the oil infrastructure layer. We clipped the GEM pipeline data to our study region using `st_intersection`, then plotted pipelines in light blue and extraction sites from GOGPT in dark blue. This map shows us the spatial footprint of oil activity before we look at how it relates to conflict.

```{r map2}
gem_clipped <- st_intersection(gem, st_union(countries))

legend_data2 <- data.frame(
  x = c(0, 0),
  y = c(0, 0),
  type = c("Extraction Sites", "Pipelines (——)")
)

ggplot() +
  geom_sf(data = countries, fill = "#f5f0e8", color = "grey40", linewidth = 0.4) +
  geom_sf(data = gem_clipped, color = "#1a6b9e", linewidth = 0.5, alpha = 0.5) +
  geom_sf(data = gogpt, color = "#0a3d62", size = 0.6, alpha = 0.8) +
  geom_point(data = legend_data2, aes(x = x, y = y, color = type), size = 0, alpha = 0) +
  geom_sf_text(data = countries, aes(label = ADMIN), size = 2.5, color = "grey30", fontface = "italic") +
  scale_color_manual(
    name   = NULL,
    values = c("Extraction Sites" = "#0a3d62", "Pipelines (——)" = "#1a6b9e"),
    guide  = guide_legend(override.aes = list(size = 3, alpha = 1))
  ) +
  coord_sf(xlim = c(34, 62), ylim = c(12, 42), expand = FALSE) +
  labs(
    title    = "Oil Infrastructure in the Middle East",
    subtitle = "Pipelines and extraction sites across eight study countries",
    caption  = "Sources: Global Energy Monitor (GEM), Global Oil & Gas Production Tracker"
  ) +
  theme_void() +
  theme(
    plot.title      = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle   = element_text(size = 10, color = "grey40", hjust = 0.5),
    plot.caption    = element_text(size = 8,  color = "grey50", hjust = 0.5),
    plot.margin     = margin(10, 10, 10, 10),
    legend.position = "left",
    legend.text     = element_text(size = 8)
  )
```



Finally, we overlaid both layers on the same map, conflict events in red on top of the oil infrastructure in blue. This combined view is our main baseline reference: it lets us visually assess whether conflict events tend to cluster near pipelines and extraction sites, which is the central question driving the rest of our analysis.

```{r map3}
legend_data3 <- data.frame(
  x = c(0, 0, 0),
  y = c(0, 0, 0),
  type = c("Conflict Events", "Extraction Sites", "Pipelines (——)")
)

ggplot() +
  geom_sf(data = countries, fill = "#f5f0e8", color = "grey40", linewidth = 0.4) +
  geom_sf(data = gem_clipped, color = "#1a6b9e", linewidth = 0.5, alpha = 0.5) +
  geom_sf(data = gogpt, color = "#0a3d62", size = 0.6, alpha = 0.8) +
  geom_sf(data = conflict, color = "#e8000d", size = 0.6, alpha = 0.5) +
  geom_point(data = legend_data3, aes(x = x, y = y, color = type), size = 0, alpha = 0) +
  geom_sf_text(data = countries, aes(label = ADMIN), size = 2.5, color = "grey30", fontface = "italic") +
  scale_color_manual(
    name   = NULL,
    values = c("Conflict Events" = "#e8000d", "Extraction Sites" = "#0a3d62", "Pipelines (——)" = "#1a6b9e"),
    guide  = guide_legend(override.aes = list(size = 3, alpha = 1))
  ) +
  coord_sf(xlim = c(34, 62), ylim = c(12, 42), expand = FALSE) +
  labs(
    title    = "Armed Conflict and Oil Infrastructure in the Middle East",
    subtitle = "Conflict events (red), extraction sites and pipelines (blue)",
    caption  = "Sources: ACLED, Global Energy Monitor (GEM), Global Oil & Gas Production Tracker"
  ) +
  theme_void() +
  theme(
    plot.title      = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle   = element_text(size = 10, color = "grey40", hjust = 0.5),
    plot.caption    = element_text(size = 8,  color = "grey50", hjust = 0.5),
    plot.margin     = margin(10, 10, 10, 10),
    legend.position = "left",
    legend.text     = element_text(size = 8)
  )
```

