---
title: "Methodology: Spatial Relationship Between Oil Infrastructure and Conflict in the Middle East"
date: "`r Sys.Date()`"
output: pdf_document
---

# Introduction

This project examines how proximity to oil infrastructure shapes the spatial distribution of armed conflict events in the Middle East. The analysis focuses on eight countries: Saudi Arabia, Iraq, Iran, Kuwait, United Arab Emirates, Qatar, Syria, and Yemen.

We use geospatial methods to:
- identify cities near oil infrastructure
- measure their proximity to conflict events
- compare patterns across different buffer distances

---

# Setup

```{r setup, message=FALSE, warning=FALSE}
library(sf)
library(dplyr)
library(stringr)
library(rnaturalearth)  

# Load data
gogpt_3857 <- readRDS("data/clean/gogpt_3857.rds")
acled <- readRDS("data/clean/acled_clean.rds")
conflict_sf <- acled$crs_3857

```

# Basemap

```{r basemap}
world <- ne_countries(scale = 50, returnclass = "sf")

countries <- c(
  "Saudi Arabia", "Iraq", "Iran",
  "Kuwait", "United Arab Emirates",
  "Qatar", "Syria", "Yemen"
)

world_subset <- world %>%
  filter(admin %in% countries)

world_subset <- st_transform(world_subset, 3857)
```

# Data preparation

## Oil infraestructure

```{r oil infraestructure}
oil_sf <- gogpt_3857 %>%
  filter(str_detect(Fuel, "oil"))

nrow(oil_sf)
unique(oil_sf$Country)

plot(st_geometry(world_subset), col = "grey90", border = "black")
plot(st_geometry(oil_sf), add = TRUE, col = "red", pch = 16)
```

# Conflict data

```{r conflict}
conflict_sf <- acled$crs_3857

### Filter time period ###
conflict_sf <- conflict_sf %>%
  filter(WEEK >= as.Date("2010-01-01"))

st_crs(conflict_sf)
unique(conflict_sf$COUNTRY)
nrow(conflict_sf)

plot(st_geometry(world_subset), col = "grey90", border = "black")
plot(st_geometry(conflict_sf), add = TRUE, pch = 16, cex = 0.5)
```

## Cities data

```{r cities}
cities_sf <- ne_download(
  scale = 10,
  type = "populated_places",
  returnclass = "sf"
)

cities_sf <- cities_sf %>%
  filter(ADM0NAME %in% countries)

cities_sf <- st_transform(cities_sf, 3857)

nrow(cities_sf)

plot(st_geometry(world_subset), col = "grey90", border = "black")
plot(st_geometry(cities_sf), add = TRUE, pch = 16, col = "blue")
```

# Methodology 

## Buffer creation and city tagging 

```{r buffer creation and city tagging}
oil_buffer_25 <- st_buffer(oil_sf, dist = 25000)
oil_buffer_50 <- st_buffer(oil_sf, dist = 50000)

cities_sf$near_oil_25km <- lengths(
  st_intersects(cities_sf, oil_buffer_25)
) > 0

cities_sf$near_oil_50km <- lengths(
  st_intersects(cities_sf, oil_buffer_50)
) > 0

table(cities_sf$near_oil_25km)
table(cities_sf$near_oil_50km)
```

## Visual check 

```{r visual check}
plot(st_geometry(world_subset), col = "grey90", border = "black")
plot(st_geometry(oil_buffer_25), col = adjustcolor("lightblue", alpha.f = 0.4), add = TRUE)
plot(st_geometry(cities_sf), add = TRUE, pch = 16)

plot(st_geometry(world_subset), col = "grey90", border = "black")
plot(st_geometry(oil_buffer_50), col = adjustcolor("lightgreen", alpha.f = 0.4), add = TRUE)
plot(st_geometry(cities_sf), add = TRUE, pch = 16)
```

## Conflict events within buffers

```{r conflict events within buffers}
conflict_in_25 <- lengths(
  st_intersects(conflict_sf, oil_buffer_25)
) > 0

conflict_in_50 <- lengths(
  st_intersects(conflict_sf, oil_buffer_50)
) > 0

table(conflict_in_25)
table(conflict_in_50)
```

## Distance to nearest conflict

```{r distance to nearest conflict}
nearest_index <- st_nearest_feature(cities_sf, conflict_sf)

nearest_dist <- st_distance(
  cities_sf,
  conflict_sf[nearest_index, ],
  by_element = TRUE
)

cities_sf$dist_to_conflict_km <- as.numeric(nearest_dist) / 1000

summary(cities_sf$dist_to_conflict_km)
```

# Results

## Mean distance to conflict (25 km)

```{r mean distance to conflict (25 km)}
cities_sf %>%
  st_drop_geometry() %>%
  group_by(near_oil_25km) %>%
  summarise(mean_distance = mean(dist_to_conflict_km, na.rm = TRUE))
```

Cities located near oil infrastructure exhibit lower average distances to conflict events compared to those located farther away.

## Mean distance to conflict (50 km)

```{r mean distance to conflict (50 km)}
cities_sf %>%
  st_drop_geometry() %>%
  group_by(near_oil_50km) %>%
  summarise(mean_distance = mean(dist_to_conflict_km, na.rm = TRUE))
```

The difference becomes more pronounced when using a larger buffer, suggesting that the spatial influence of oil infrastructure extends beyond immediate proximity.
